---
title: "Life Expectancy"
author: "Amy Herring"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

## Introduction

How has life expectancy from infancy changed over time around the world?

### Packages

We will use the __tidyverse__, __lubridate__, and __scales__ packages for the data wrangling and visualization, and the __DT__ package for interactive display of tabular output.

```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(DT)
```

### Data

The data we're using originally come from the Institute for Health Metrics and Evaluation.

```{r load-data}
life <- readr::read_csv("lifeexp/lifeexpectancy_infant.csv")
```


## Life Expectancy {#lifeexp}

Let's  create a data visualization that displays how life expectancy in the US has changed over time for females, comparing it to life expectancy in India and Iran.

We can easily change which countries are being plotted by changing which countries the code above `filter`s for. Note that the country name should be spelled and capitalized exactly the same way as it appears in the data. See the [Appendix](#appendix) for a list of the countries in the data.

```{r plotlifeexp, fig.width=10, fig.height=6, message=FALSE}
life %>%
  filter(sex == "Female") %>%
  filter(location %in% c("United States of America", "Iran (Islamic Republic of)","India")) %>%
  ggplot(mapping = aes(x = year, y = lifeexp, color = location)) +
  geom_point() +
  labs(
    title = "Female Life Expectancy Over Time",
    subtitle = "1990 to 2019",
    y = "Life Expectancy (Years)",
    x = "Year",
    color = "Country"
  )
```



## Appendix {#appendix}

Below is a list of countries in the dataset:

```{r list-countries, echo=FALSE}
life %>% 
  select(location) %>%
  arrange(location) %>% 
  distinct() %>%
  datatable()
```
