---
title: "Welcome to STA 198 Labs!"
date: "09.06.2021"
output:
  xaringan::moon_reader:
    css: 
      - css/xaringan-themer.css
      - css/slides.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## What to expect in lab 

- Introduction to the lab assignment (~ 5 - 10 minutes)

- Work on the lab assignment (individual today, but generally done in teams)

- Lab instructions posted in the [schedule](https://sta198f2021.github.io/website/#schedule) for each week on the course website.

- Start each lab by finding your assignment repo in the [GitHub course organization](https://github.com/STA198F2021)
  - This is where you will find the R Markdown document and data to get started
  
---


## Check: Do you have the lab-02 repo?

- Go to the GitHub course organization: https://github.com/STA198F2021

- You should see a repo with the prefix **lab-02-**

- If you do not have this repo, please let your TAs know! 

---

class: middle, center

# Demos

---

## Demos 

**Follow along as your TA demonstrates the following (next week we'll assume you've got it down!)** 

- Access RStudio using the class container
- Configure Git using SSH 
- Clone RStudio repo and start new project 
- Knit document and produce PDF 
- Update name in YAML 
  - Knit, commit, push changes to GitHub 
  - See updates in your GitHub repo
  
---

## Tips for today

1. Don't worry about the data wrangling code; we'll get to that later in the course.

2. [Learn more about ggplot2 color options!](http://sape.inf.usi.ch/quick-reference/ggplot2/colour)

---

class: middle 

## You're now ready to complete the rest of lab!


  