---
title: "Discrete Probability Distributions"
subtitle: "<br><br> Introduction to Global Health Data Science"
author: "[Back to Website](https://sta198f2021.github.io/website/)"
date: "<br> Prof. Amy Herring"
output:
  xaringan::moon_reader:
    css: 
      - css/xaringan-themer.css
      - css/slides.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

```

class: middle

# Probability Distributions

---

### Probability Distributions

When we learned how to find probabilities using basic principles, we focused on one particular outcome or event, like the probability of being a smoker. We now explore the bigger picture by exploring all possible values of a discrete random variable, along with their associated probabilities. This list of possible values and probabilities is called the *probability distribution* of the random variable.

---

### Why do we care about probability distributions?

We need probability distributions to make statements about how
likely an event may be. For example, suppose we are interested in
comparing two groups of pancreatic cancer patients, with one
group on the standard treatment and one group on an experimental
new treatment. The statistical test we use to compare the two
groups depends on the variable of interest $X$ and its probability
distribution. One of you could define $X$ to be a 0/1 variable
indicating whether each patient survives 1 year, and someone else
could instead define X as the number of months a patient survives.
In these two cases, each definition of $X$ has a different probability distribution, and you would use a different type of test for each case.

---

### Example: Y Chromosomes

Consider Y chromosomes in the US population.  51.2% of births are to babies with Y chromosomes, and 48.8% of babies do not have Y chromosomes. Thus the probability of having a baby with a Y chromosome is 0.512.

- The probability of an event is $\geq 0$ and $\leq 1$.

- The sum of the probabilities of all possible events is exactly 1 (e.g., 0.512+0.488=1).

---

### Bernoulli Distribution

Consider a dichotomous (two-level) random variable $Y$. By definition, $Y$ must assume one of two possible values, e.g.

- failure or success
- dead or alive
- Duke student or not
- current smoker or not
- heads or tails (coin flip)

A random variable of this type is known as a *Bernoulli* random variable, and we describe the probability of response using the parameter $\pi$. 

The distribution is named for [Jacob Bernoulli](https://en.wikipedia.org/wiki/Jacob_Bernoulli), who was part of a famous family of mathematicians in Switzerland.

---

### Bernoulli random variable

These variables are often coded so that $Y=1$ is called an *event* or *success*, $Y=0$ is called a *failure*, and $\pi$ is defined as the probability of a success. (This is not required, but you need to remember which level maps to $\pi$, and the other level then has probability $1-\pi$.)

Examples:
- Coin flip: let $Y=1$ if heads and $Y=0$ if tails, then $P(Y=1)=\pi=0.5$
- Vegetarian in US: $Y=1$ if vegetarian and $Y=0$ if not, then $P(Y=1)=\pi=0.05$ and $P(Y=0)=1-\pi=1-0.05=0.95$
- Vegetarian in India: $Y=1$ if vegetarian and $Y=0$ if not, then $P(Y=1)=\pi=0.31$ and $P(Y=0)=1-\pi=1-0.31=0.69$

---

### Bernoulli distribution

- $Y$ takes value 1 with probability $\pi$ and value 0 with probability $1-\pi$
- $P(Y=y)=\pi^y(1-\pi)^{1-y}$
  - $P(Y=1)=\pi^1(1-\pi)^0=\pi$ as $x^0=1$ for any $x$
  - $P(Y=0)=\pi^0(1-\pi)^1=1-\pi$
- We don't really need all this formality as it's probably easier just to keep track of $\pi$ and $1-\pi$
- However, we want to extend this to a more complex setting -- for example, in a randomly-selected group of 3 high school students, how surprising would it be to get 2 who have smoked e-cigarettes in the past month?

---

### Case Study: E-Cigarettes

- The [CDC reports](https://www.cdc.gov/tobacco/data_statistics/fact_sheets/youth_data/tobacco_use/index.htm) that 19.6% of high school students have smoked e-cigarettes in the past 30 days. We'll round this to 20% for simplicity.
- $P(Y=1)=P(Smoker)=\pi=0.2$ and $P(Y=0)=0.8$
- Now suppose we randomly select two independent high school students and define a new random variable $X$ representing the number of smokers. X can take the values 0, 1, or 2. Let $Y_1$ be the smoking status of the first student ($Y_1=1$ if the student smokes) and $Y_2$ be the smoking status of the second student.

.pull-left[
Next we'll talk about how to get the *probability distribution* of $X$.

]
.pull-right[
| $Y_1$ | $Y_2$ | $X$ | $P(X)$ |
|:----:|:----:|:----:|:----:|
| 0 | 0 | 0 | |
| 1 | 0 | 1 | |
| 0 | 1 | 1 | |
| 1 | 1 | 1 | |
]

---

Let $A_1$ be the event that $Y_1=1$ and let $A_2$ be the event that $Y_2=1$. 

Because the people are independent, then

$$\begin{array}{ll}
P(both smoke) & = P(A_1 \cup A_2) \\
& = P(A_1) P(A_2 \mid A_1) \\
& = P(A_1) P(A_2) \\
& = \pi \times \pi = 0.2(0.2)=0.04.$$






